---
output:
  pdf_document: default
  html_document: default
---
# Forest Carbon Codefest data skills example
## Create a graphic with statistics from Monitoring Trends in Burn Severity (MTBS) perimeters

CU Boulder CIRES Earth Lab

A basic level of coding ability will be required to attend the Forest Carbon Codefest, such that any given participant can (in the coding language of their choosing): i) open and plot tabular and geospatial data, ii) compute summary statistics or create basic statistical models, and iii) create simple data visualizations.

Not sure what that means? This markdown contains a few example graphics; you should be able to create something similar, if given the appropriate data. Note that while this example is in R, similar ability in any coding language is sufficient!

## SETUP

```{r, warning = FALSE, results = 'hide', message = FALSE}

#Check the required libraries and download if needed
list.of.packages <- c("tidyverse", 
                      "terra",
                      "sf",
                      "tmap",
                      "here",
                      "mblm", #Median-based linear models (i.e. thiel-sen)
                      "tigris", #US data
                      "scales", #add commas to ggplot axis
                      "tinytex") 
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

invisible(lapply(list.of.packages, library, character.only = TRUE)) #apply library function to all packages without producing any output

```

## Read in MTBS & manipulate

```{r, message=FALSE, warning = FALSE, results = 'hide'}

#Read in MTBS data, ensure valid, add burn year
mtbsPerims <- sf::st_read("C:/Users/tyler/OneDrive - UCB-O365/dev/fast-fires/data/mtbs_perimeter_data_1984_2021/mtbs_perims_DD.shp") %>% 
  st_transform(st_crs(4269)) #Data from https://www.mtbs.gov/


head(mtbsPerims)

#Get burn year for each polygon and the burn area in hectares instead of acres
mtbsPerims <- mtbsPerims %>%
  dplyr::mutate(Burn_Yr = year(as.Date(Ig_Date))) %>%
  dplyr::mutate(BurnHa = BurnBndAc * 0.404686) %>%
  sf::st_make_valid()

#Ensure polygon validity
sf::st_is_valid(mtbsPerims) %>% unique()


#Get shapefiles for the US
# Fetch all US states
usa <- tigris::states() %>% 
  st_transform(st_crs(4269))

# Filter for contiguous states (excluding Alaska and Hawaii)
conus <- usa[usa$STUSPS %in% c("WA", "ID", "MT", "ND", "MN", "MI", "ME", "WI", "VT", "NH", "NY", "SD", "WY", "OR", "IA", "NE", "IL", "NV", "UT", "CO", "KS", "OK", "NM", "AZ", "AR", "MO", "MS", "AL", "GA", "SC", "NC", "TN", "KY", "IN", "OH", "WV", "VA", "PA", "MD", "DE", "NJ", "CT", "RI", "MA", "CT", "CA", "TX", "FL", "LA"),]

west <- usa[usa$STUSPS %in% c("WA", "OR", "CA", "ID", "MT", "WY", "NV", "AZ", "CO", "NM", "UT"),]


#Filter perimeters to just include the area we want
mtbsPerimsWest <- sf::st_filter(mtbsPerims, west)

#Plot the perimeters (could use mapview library here if wanted an interactive visual)
tmap::tm_shape(mtbsPerimsWest) +
  tmap::tm_fill(col = "Burn_Yr") +
  tmap::tm_layout(main.title = "MTBS Perimeters in the \nWestern U.S. by Burn Year")

```

\newpage

## Create summaries of MTBS data and plot

```{r}

#summarize burned area by hectares
get.wildfire.burn.ha.summary <- function(perims) {
  burnedAc <- perims %>%
    filter(Incid_Type == "Wildfire") %>%
    select(Burn_Yr, BurnHa) %>%
    as.data.frame() %>%
    group_by(Burn_Yr) %>%
    summarise(totBurnHa = sum(BurnHa))
  return(burnedAc)
}

#Get summary stats
westBurned <- get.wildfire.burn.ha.summary(mtbsPerimsWest)

#Get theil-sen fits and graph
theilSenFitWest <- mblm::mblm(totBurnHa ~ Burn_Yr, westBurned, repeated = FALSE)

#Plot and add Theil-Sen fit line to graph
ggplot(westBurned, aes(x = Burn_Yr, y = totBurnHa)) +
  geom_col(fill = "maroon") +
  labs(title = "Western U.S. Burned Area 1984-2021") +
  xlab("Year") +
  ylab("Total area burned by wildfire (ha)") +
  scale_y_continuous(labels=comma) +
  geom_abline(intercept = theilSenFitWest$coefficients["(Intercept)"],
              slope = theilSenFitWest$coefficients["Burn_Yr"],
              linetype = "dashed",
              linewidth = 0.8) +
  theme_light()


```

## Estimator coefficients and significance


```{r, warning=FALSE}
#Get estimator significance
summary.mblm(theilSenFitWest)
```


